---
title: "assignment"
output: html_document
date: "2026-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=FALSE, warning = FALSE,  message = FALSE)
```

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)      # read alignment
  library(rtracklayer)   # read/write genomic file formats
  library(Biostrings)    # handling sequences
  library(Rfastp)        # read QC/trimming
  library(epiwraps)
})
ah <- AnnotationHub()
```

```{r}
setwd("/Users/marco/Desktop/ethEpigen2026_material/week03")
```

#### Downloading fastq file of CTCF
```{r}
options(timeout=3600)
download.file("https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz","ENCFF127RRR.fastq.gz", mode="wb")
```

```{bash engine.opts='-l', eval=FALSE}
mkdir -p raw_fastqc
fastqc -o raw_fastqc --threads 4 *.fastq.gz
```

```{bash engine.opts='-l', eval=FALSE}
multiqc -o raw_fastqc raw_fastqc
```

```{bash, eval=FALSE}
export PATH="/Users/marco/miniconda3/bin:$PATH"

mkdir -p trimmed

trimmomatic SE -threads 4 \
-summary trimmed/input.stats \
-phred33 \
ENCFF127RRR.fastq.gz \
trimmed/trimmed.fastq.gz \
ILLUMINACLIP:/Users/marco/miniconda3/share/trimmomatic-0.40-0/adapters/TruSeq3-SE.fa:2:15:4:4:true \
LEADING:20 \
TRAILING:20 \
SLIDINGWINDOW:4:15 \
MINLEN:25

#Surviving 95% reads
```
 
#### Loading the genome
```{r}
options(timeout=5000)
# we get the genome sequence from AnnotationHub
genome <- ah[["AH49674"]]
# we create a new directory that will contain the genome index
dir.create("BDGP6_genome")
# we write the genome sequence in fasta format
export(import.2bit(genome), "BDGP6_genome/genome.fasta.gz", compress=TRUE)
```

#### Building a bowtie2 index
```{bash engine.opts='-l'}
bowtie2-build --threads 4 BDGP6_genome/genome.fasta.gz BDGP6_genome/bowtie2
```

#### Alignment
```{bash engine.opts='-l'}
mkdir -p aligned

(bowtie2 -p 4 -x BDGP6_genome/bowtie2 -U trimmed/trimmed.fastq.gz) 2> aligned/aligned.bowtie2 |\
  samtools view -bS - | samtools sort -@4 -m 2G - > aligned/aligned.bam
samtools index aligned/aligned.bam

#    3693508 (100.00%) were unpaired; of these:
#    164910 (4.46%) aligned 0 times
#    2910891 (78.81%) aligned exactly 1 time
#    617707 (16.72%) aligned >1 times
#.   95.54% overall alignment rate
```

#### Peak calling using MACS
```{bash engine.opts='-l'}
# hs mm dm
mkdir -p peaks
macs3 callpeak --outdir peaks -n ctcf --gsize dm -t aligned/aligned.bam -B #-c aligned/input.bam don't have + #.bdg file for coverage
```

```{r, eval=TRUE}
peaks <- rtracklayer::import("peaks/ctcf_peaks.narrowPeak")
```

```{r, echo = FALSE}
library(Gviz)
library(rtracklayer)
#loading files
peaks <- rtracklayer::import("peaks/ctcf_peaks.narrowPeak")
treat_gr <- rtracklayer::import("peaks/ctcf_treat_pileup.bdg", format = "bedGraph")
lambda_gr <- rtracklayer::import("peaks/ctcf_control_lambda.bdg", format = "bedGraph")
xls <- read.table("peaks/ctcf_peaks.xls", header = TRUE, comment.char = "#")

#select the peak
i <- 2000
#selecting subset of peaks (Global vs Gene (protein_coding))
my_peak <- peaks[i] 

#---------------------------------------------------------------------------------------------------------------#
#Matching narrowPeak and XLS (handles the 1bp offset)
match_idx <- which(abs(xls$start - start(my_peak)) <= 1 & 
                   (xls$chr == as.character(seqnames(my_peak)) | 
                    xls$chr == gsub("chr", "", as.character(seqnames(my_peak)))))[1]

if(is.na(match_idx)) stop(paste("No match", i))

dati_xls <- xls[match_idx, ]

#extract values for labels
pileup_val   <- as.numeric(dati_xls$pileup)
fc_effective <- as.numeric(dati_xls$fold_enrichment)
bg_colored <- pileup_val / fc_effective

#track signal
strack <- DataTrack(range = treat_gr, 
                    name = "Signal", 
                    type = "histogram", 
                    fill = "gray70", 
                    col = "transparent")

#background line
ltrack <- DataTrack(range = lambda_gr, 
                    name = "Background", 
                    type = "polygon",
                    type = "l", 
                    col = "red", 
                    lwd = 3) 

#overlay
otrack <- OverlayTrack(trackList = list(strack, ltrack))

#peaks track
labels <- rep(" ", length(peaks)); labels[i] <- "TARGET"
colors <- rep("royalblue", length(peaks)); colors[i] <- "red"
ptrack <- AnnotationTrack(peaks, name = "Peaks", id = labels, showId = TRUE,
                          fill = colors, col = "transparent", just.group = "above")

#final plot
plotTracks(list(otrack, ptrack),
           chromosome = as.character(seqnames(my_peak)),
           from = start(my_peak) - 10000, 
           to = end(my_peak) + 10000,
           main = paste0("Peak ", i, " | FC: ", round(fc_effective, 2), "\n",
                         "Local Background (Avg): ", round(bg_colored, 2), "reads"),
           cex.main = 0.9,
           fontface.main = 2,
           ylim = c(0, max(pileup_val * 1.5, 50)))
#---------------------------------------------------------------------------------------------------------------#
```

#### Filter the peak that fall into a gene
```{r}
library(ensembldb)
library(GenomicRanges)

#Save the original index on peaks
peaks$old_index <- seq_along(peaks)

#Get the Drosophila EnsDb and extract protein-coding genes
ensdb_dro <- ah[["AH119285"]]

prot_coding = FALSE

if(prot_coding) {

  dro_pcg <- genes(ensdb_dro, filter = ~ gene_biotype == "protein_coding")
  
  #Filter the peaks that fall inside a protein coding gene
  peaks_filtered <- subsetByOverlaps(peaks, dro_pcg)

} else {
  dro_all <- genes(ensdb_dro)
  
  #Filter the peaks that fall inside a gene
  peaks_filtered <- subsetByOverlaps(peaks, dro_all)
}
save(peaks_filtered, file = "peaks_filtered.RData")
```

## PLOT
```{r, eval=TRUE,fig.width=12, fig.height=8, out.width="100%", fig.align='center'}
library(Gviz)
library(rtracklayer)

#LOADING FOR KNIT
options(ucscChromosomeNames = FALSE)
library(AnnotationHub)
ah <- AnnotationHub()
ensdb_dro <- ah[["AH119285"]]

#loading files
load("peaks_filtered.RData")
peaks <- rtracklayer::import("peaks/ctcf_peaks.narrowPeak")
treat_gr <- rtracklayer::import("peaks/ctcf_treat_pileup.bdg", format = "bedGraph")
lambda_gr <- rtracklayer::import("peaks/ctcf_control_lambda.bdg", format = "bedGraph")
xls <- read.table("peaks/ctcf_peaks.xls", header = TRUE, comment.char = "#")

#select the peak
i <- 1500
#selecting subset of peaks (Global vs Gene (protein_coding))
#my_peak <- peaks[i] 
my_peak <- peaks_filtered[i]

#Matching narrowPeak and XLS (handles the 1bp offset)
match_idx <- which(abs(xls$start - start(my_peak)) <= 1 & 
                   (xls$chr == as.character(seqnames(my_peak)) | 
                    xls$chr == gsub("chr", "", as.character(seqnames(my_peak)))))[1]

if(is.na(match_idx)) stop(paste("No match", i))

dati_xls <- xls[match_idx, ]

#extract values for labels
pileup_val   <- as.numeric(dati_xls$pileup)
fc_effective <- as.numeric(dati_xls$fold_enrichment)
bg_colored <- pileup_val / fc_effective

#track signal
strack <- DataTrack(range = treat_gr, 
                    name = "Signal", 
                    type = "histogram", 
                    fill = "gray70", 
                    col = "transparent")

#background line
ltrack <- DataTrack(range = lambda_gr, 
                    name = "Background", 
                    type = "polygon",
                    type = "l", 
                    col = "red", 
                    lwd = 3) 

#overlay
otrack <- OverlayTrack(trackList = list(strack, ltrack))

#peaks track
labels <- rep(" ", length(peaks)); labels[i] <- "TARGET"
colors <- rep("royalblue", length(peaks)); colors[i] <- "red"
ptrack <- AnnotationTrack(peaks, name = "Peaks", id = labels, showId = TRUE,
                          fill = colors, col = "transparent", just.group = "above")


#Add a Genome Axis (improves styling and readability)
atrack <- GenomeAxisTrack(add53 = TRUE, add35 = TRUE)

## ADDING SYMBOL LABEL INSTEAD OF TRANSCRIPT ID
#Extraction of data in the peak region
gene_data <- ensembldb::getGeneRegionTrackForGviz(ensdb_dro, 
                                        chromosome = as.character(seqnames(my_peak)), 
                                        start = start(my_peak) - 10000, 
                                        end = end(my_peak) + 10000)

#Manual substitution of Symbol instead of Transcript ID
gene_data$transcript <- ifelse(!is.na(gene_data$symbol) & gene_data$symbol != "", 
                               gene_data$symbol, 
                               gene_data$gene)

#track with symbol and not ID
gtrack <- GeneRegionTrack(gene_data,
                          name = "Genes",
                          transcriptAnnotation = "transcript", 
                          collapseTranscripts = "longest",     
                          fill = "#8282d2", 
                          col = "transparent")

plotTracks(list(atrack, otrack, ptrack, gtrack),
           chromosome = as.character(seqnames(my_peak)),
           from = start(my_peak) - 10000, 
           to = end(my_peak) + 10000,
           showId = TRUE,           
           geneSymbol = TRUE,      
           main = paste0("Peak n. ", peaks_filtered$old_index[i], " | Fold Change: ", round(fc_effective, 2), "\n",
                         "Local Background (Avg): ", round(bg_colored, 2), " reads"),
           cex.main = 0.9,
           fontface.main = 2,
           ylim = c(0, max(pileup_val * 1.5, 50)))
```

#### Fold Change: the signal enrichment relative to the estimated local background model
#### Background (Red line): the expected noise level calculated by MACS based on the local genomic context

## REPORT
#### How many reads (and what percentage) were mapped
```{bash engine.opts='-l', eval=TRUE}
samtools flagstat aligned/aligned.bam | grep "mapped ("
```
#### How many peaks were found
```{r, eval=TRUE}
length(peaks)
```


